Copulas and their potential for ecology

### 2019-06-16 Paper.Rmd
further edits, increasingly minor

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have copula structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
dependence distinct from that of a normal distribution, and in particular
do they show asymmetric tail dependence? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. Such 
approaches do not utilize the power of copulas, so Q1 asks whether ecological
data contain dependency information distinct from what can be modelled with
standard approaches. (Q2) If the answer to
Q1 is "yes," what are some possible causes/mechanisms of non-normal
copula structure? And (Q3) what are the consequences of non-normal copula
structure and tail dependence for our understanding of ecology and for 
applications? To address Q1 (the second through fourth sections of the paper), 
we analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
To address Q2 (the fifth and sixth sections of the paper) we use 
simple models. To address Q3 (the seventh and eigths sections) we use 
both data and models. The ninth section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, non-normal copulas have not been used in ecology. 
We argue that benefits to ecology of using a complete 
copula description of dependence will be substantial.

@@

Datasets included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data (Table \ref{tab_data_info}), selected to span multiple fields and levels
of organization within ecology. Copula structure describing 
the relationship between atmospheric weather variables such as rainfall or wind speed 
as measured in multiple locations has been 
examined previously in the meteorological literature [e.g., @Serinaldi2008; @li2013]. We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database [RaCA; @Wills2014].
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (Mg C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for sampling locations).
Species-trait data were average species basal metabolic rate (BMR, KJ per hour) and 
body mass (grams) for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey (RIS) suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years. 
Our second population dataset comprised average annual plankton abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. 
Community-level data, obtained from the Cedar Creek Ecosystem Science Reserve, were 
plant aboveground biomass [@ccdataple120] and Shannon's diversity index (computed from plant 
species percent cover data [@ccdatapce120]) for `r nrow(d_cc[[5]])` plots [@ccdataplotse120], 
each 9m by 9m, as sampled in the years 1996-2000 and 2007, each year 
analyzed separately [@Tilman2001; @Tilman2006]. 
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. See Appendix \ref{SM-Data_details} for additional data details.

@@

<!--summary table with info of all data used in Paper-->
```{r tab_summary_data_info, echo=F, results="asis",message=F}
library(kableExtra)
library(dplyr)
tab_data_info<-as.data.frame(matrix(NA,nrow=8,ncol=4))
colnames(tab_data_info)<-c("Data","Category","No. of measurements we used", "References")
tab_data_info$Data<-c("Soil C and N","Bird body masses and BMR", "Mammal body masses and BMR", 
                      "Green spruce and other aphid species abundances", "Leaf-curling plum and other aphid first flight dates","\\textit{Ceratium furca} and other plankton taxa abundances", "Plant diversity and aboveground biomass","Methane-flux")
tab_data_info$Category<-c("Environmental","Species-trait","Species-trait",
                          "Population","Phenological","Population","Community level","Ecosystem functioning")
tab_data_info$`No. of measurements we used`<-c(paste(nrow(d_soilCN)," locations",sep=""),
                                               paste(nrow(d_birdsBMR)," birds",sep=""),
                                               paste(nrow(d_mammalsBMR)," mammals",sep=""),
                                               "10 locations with at least 30-yr. timeseries for each 20 sp.",
                                               "10 locations with at least 30-yr. timeseries for each 20 sp.",
                                               "14 locations with at least 45-yr. timeseries for each 22 taxa",
                                               paste(nrow(d_cc[[5]])," plots",sep=""),
                                               "13 locations with at least 50 dates of data")
tab_data_info$References<-c("US Rapid Carbon Assessment database (RaCA), Wills et al. (2014)", "McNab (2009)","McNab (2008)",
                            "Rothamsted Insect Survey","Rothamsted Insect Survey","Continuous Plankton Recorder Survey",
                            "Cedar Creek Ecosystem Science Reserve, biodiversity experiment e120",
                            "Great Miami Wetland Mitigation Bank, Smyth et al. (2019)")

@@

In addition to our model selection approach, we also used 
nonparametric approaches to provide greater 
confidence in our answer to Q1. We used three 
statistics which quantify the extent to which 
the normalized ranks $u_i$ and $v_i$ are 
related in any part of their distributions; we here describe the 
statistics conceptually, with details in Appendix \ref{SM-nonparam_stats}.
The statistics are defined with positively associated
variables in mind. All our variables were positively associated
when they were significantly associated.
Given two bounds 
$0 \leq l_b < u_b \leq 1$, define the lines $u+v=2l_b$ and 
$u+v=2u_b$, which intersect the unit square (Figs \ref{fig_cop_pedag1}H, 
\ref{SM-fig_stat_image}). 
Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and $u_b\leq 0.5$ gives information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ gives information about dependence in 
the right parts. The first statistic, 
$\cor_{l_b,u_b}(u,v)$, is the portion of the Spearman 
correlation of $u_i$ and $v_i$ that is attributable to the points 
between the bounds.
Larger values indicate stronger positive association.
We also defined a statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar 
interpretation to $\cor_{l_{b},u_{b}}(u,v)$.
Our third statistic, $\Dsq_{l_b,u_b}(u,v)$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, for which large values inidicate strong dependence
between the bounds, small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6.5in]{./Results/BivarDataPlot.pdf}
\caption{Upper panels show raw data plots for (A) $\log_{10}$(soil C) and $\log_{10}$(soil N) data, (B) $\log_{10}$(basal metabolic rate (BMR)) vs. $\log_{10}$(body mass) 
for birds, (C) the same for mammals, and (D) above-ground plant biomass vs. Shannon's index (\textit{H}) from Cedar Creek. Bottom panels show the corresponding copula plots. See Data for units used in upper panels. \label{fig_biv_multi_raw_cop}}
\end{center}
\end{figure}

@@

<!--bivariate example-->
For the soil C and N data (Data, Table \ref{tab_data_info}, 
Fig. \ref{fig_biv_multi_raw_cop}A,E), variables were 
non-independent ($p=$ `r BivMS_res_Loecke$IndepTestRes`, to within the precision available 
from `BiCopIndTest` in the `VineCopula` package); non-independence is also visually apparent.
The Kendall correlation was `r round(BivMS_res_Loecke$TauVal,2)`. We fitted our 16 copulas
to the data and computed AIC and BIC weights and corresponding
lower- and upper-tail statistics for the fitted distributions (Table \ref{SM-tab_soilCNfit}).
`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]` was the 
minimum-AIC copula, with AIC weight `r round(max(BivMS_res_Loecke$InfCritRes$AICw),2)`, whereas the 
N copula had AIC weight essentially 0, answering the 
first part of Q1 for these data. 
Model-averaged lower- and upper-tail dependence statistics were `r round(BivMS_res_Loecke$relLTdep_AICw,3)` 
and `r round(BivMS_res_Loecke$relUTdep_AICw,3)`, respectively (AIC weights were used for averaging). 
These values are distinct from what an N copula would give, namely, 0, answering the second part 
Q2 for these data. The values also differed substantially 
from each other, answering the third part of Q1. 
The numbers reflect stronger upper- than lower-tail
dependence, and this is also visible in the extreme upper-right
and lower-left corners of the copula plot (Fig. \ref{fig_biv_multi_raw_cop}E).

Although these results convincingly show non-N copula structure, 
we caution that the model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula 
(`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]`) 
was a poor fit ($p=$ `r BivMS_res_Loecke$GofRes_CvM` and `r BivMS_res_Loecke$GofRes_KS` for the 
Cramer-von Mises and Kolmogorov-Smirnov goodness-of-fit tests, respectively, to within the precision 
available from `BiCopGofTest`). However, nonparametric statistics verified
both that tail dependence was distinct from that of an N copula, and that it 
was asymmetric (Appendices \ref{SM-Q1bivarnonparam}, \ref{SM-surrog_test}, 
Fig. \ref{SM-fig_soilCN_nonparam}, Table \ref{SM-tab_soilCN_asym}). The 
affirmative, empirically based answer to Q1 for the soil C and N data is 
represented in Fig. \ref{MasterFigure} as a solid outline around "soil C and N"
in the right-most box in the middle row.

<!--bivariate summary-->
Table \ref{tab_bivar_summary} summarizes Q1 results 
for the bivariate datasets. The two 
variables were significantly related for all datasets 
(Table \ref{tab_bivar_summary}, row 1). A non-N copula 
always emerged as the lowest-AIC (highest-AIC-weight) copula 
(Table \ref{tab_bivar_summary}, row 2). 
The N copula was always a poor fit compared to the highest-AIC-weight copula 
(Table \ref{tab_bivar_summary}, rows 3-4), answering the first
part of Q1. 
Lower- and upper-tail dependence statistics often differed substantially from 0 
(Table \ref{tab_bivar_summary}, rows 7-8) and from each other (Table \ref{tab_bivar_summary}, row 9), 
answering the second and third parts of Q1. 
The first three datasets had stronger upper- 
than lower-tail dependence, and the last dataset had the reverse, though weakly. 
Although these results convincingly show non-N copula structure, the 
model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula was a poor fit in some cases 
(Table \ref{tab_bivar_summary}, rows 5-6).
But our nonparametric statistics confirmed the same conclusions 
(Appendices \ref{SM-Q1bivarnonparam}, \ref{SM-surrog_test}).
The 
affirmative, empirically based answer to Q1 for the bivariate datasets is 
represented in Fig. \ref{MasterFigure} as solid outlines around the names of those
datasets in the middle row of boxes.
We did the same analysis 
with Cedar Creek data from other available years, 1996-1999 and 2007. For 1996 and 
2007, biomass and Shannon's index were independent; for 1997-1999, results 
were similar to 2000: the best-fitting copula was not N, and there was slightly stronger lower- 
than upper-tail dependence. Detailed results for each dataset are in 
Tables \ref{SM-tab_soilCNfit} - \ref{SM-tab_cc_asym_2000} and 
Figs. \ref{SM-fig_soilCN_nonparam} - \ref{SM-fig_cc2000_nonparam}.

We present green spruce aphid abundance (Table \ref{tab_data_info}) results
as an example multivariate analysis.
Independence was rejected for each pair of the 10 sampling locations. 
Best-fitting (lowest-AIC) copulas were non-N
for the large majority of location pairs (Table \ref{SM-tab_count_bestcop}), and 
AIC weights for best-fitting copulas (Table \ref{SM-tab_count_AICw}) 
were typically substantially 
larger than N-copula weights (Table \ref{SM-tab_count_AICw_normal}), answering the first part of Q1. Goodness of fit tests 
(Tables \ref{SM-tab_count_p_CvM}, \ref{SM-tab_count_p_KS}) in every case 
failed to reject the hypothesis that the AIC-best copula family was also 
an objectively adequate description of 
the data; the collection of 
copula families we used was sufficiently broad to characterize these data.
Model-averaged lower- (Table \ref{SM-tab_count_LTdep_AICw}) and upper-tail
(Table \ref{SM-tab_count_UTdep_AICw}) dependence statistics were exclusively 
positive and were frequently large, thereby differing 
from what an N copula would give (i.e., 0), and answering the second
part of Q1. These data have greater tail dependence 
(lower and upper) than would be expected from an N-copula null hypothesis. 
Model-averaged lower- minus upper-tail dependence statistics 
(Table \ref{SM-tab_count_LTmUT}) were positive 
for all but a few location pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`). Thus the spatial synchrony of rarity in the green spruce aphid is stronger 
than the spatial synchrony of outbreaks. This also answers the third part of Q1.
Nonparametric statistics verified that tail dependence was asymmetric for the green 
spruce aphid abundance data (Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}).

<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes Q1 results for
multivariate data. Results supported the 
conclusions that non-N copula structure, non-N tail dependence, and asymmetric tail 
dependence were common, answering Q1 in the affirmative for these data. 
Most location pairs were non-independent 
(Table \ref{tab_multivar_summary}, row 2). The large majority of non-independent location 
pairs had best-fitting copulas that were not the N copula (Table \ref{tab_multivar_summary}, 
row 3), and AIC weights of best-fitting copulas were 
better than those of N copulas (Table \ref{tab_multivar_summary}, rows 4-5).
Best-fitting copulas were nearly always considered an adequate fit 
(Table \ref{tab_multivar_summary}, row 6). Some of the datasets (green spruce aphid
abundance, \textit{Ceratium furca} abundance, methane-flux) showed stronger 
or marginally stronger lower- than upper-tail dependence 
(Table \ref{tab_multivar_summary}, rows 7-9), whereas leaf-curling plum aphid first flight
data showed the reverse. Asymmetry results were verified by nonparametric approaches 
<!--(e.g., Table \ref{tab_multivar_summary}, row 10)-->
(Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}). Detailed results are in 
Tables \ref{SM-tab_count_bestcop}-\ref{SM-tab_methane_resamp} and 
Figs \ref{SM-fig_aphid_count}-\ref{SM-fig_methane}.
The 
affirmative, empirically based answer to Q1 for the multivariate datasets is 
represented in Fig. \ref{MasterFigure} as solid outlines around the names
of those datasets in the middle row of boxes.
We also carried out the same analyses for abundance and first-flight data for the 18 
aphid species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other than the
\textit{Ceratium furca} (results not shown); conclusions supported that 
non-N copula structure and tail dependence and asymmetry of tail dependence are 
common.

@@

\noindent \textbf{Figure 3.} Upper panels show raw data plots for (A) $\log_{10}$(soil C) and $\log_{10}$(soil N) data, (B) $\log_{10}$(basal metabolic rate (BMR)) vs. $\log_{10}$(body mass) 
for birds, (C) the same for mammals, and (D) above-ground plant biomass vs. Shannon's index (\textit{H}) from Cedar Creek. Bottom panels show the corresponding copula plots. See Data for units used in upper panels.

>>>>

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have copula structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
dependence distinct from that of a normal distribution, and in particular
do they show asymmetric tail dependence? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. Such 
approaches do not utilize the power of copulas, so Q1 asks whether ecological
data contain dependency information distinct from what can be modelled with
standard approaches. (Q2) If the answer to
Q1 is "yes," what are some possible causes/mechanisms of non-normal
copula structure and tail dependence? And (Q3) what are the consequences of non-normal copula
structure and tail dependence for our understanding of ecology and for 
applications? We address Q1, in the second through fourth sections of the paper, 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2, in the fifth and sixth sections of the paper, using 
simple models. We address Q3, in the seventh and eigths sections of the paper, using 
both data and models. The ninth section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, non-normal copulas have not been used in ecology. 
We argue that benefits to ecology of using a complete 
copula description of dependence will be substantial.

@@

Datasets included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data (Table \ref{tab_data_info}), selected to span multiple fields and levels
of organization within ecology. Copula structure describing 
the relationship between atmospheric weather variables such as rainfall or wind speed, 
as measured in multiple locations through time, has been 
examined previously in the meteorological literature [e.g., @Serinaldi2008; @li2013]. We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database [RaCA; @Wills2014].
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (Mg C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for sampling locations).
Species-trait data were average species basal metabolic rate (BMR, KJ per hour) and 
body mass (grams) for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey (RIS) suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years. 
Our second population dataset comprised average annual plankton abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. 
Community-level data, obtained from the Cedar Creek Ecosystem Science Reserve, were 
plant aboveground biomass [@ccdataple120] and Shannon's diversity index (computed from plant 
species percent cover data [@ccdatapce120]) for `r nrow(d_cc[[5]])` plots [@ccdataplotse120], 
each 9m by 9m, as sampled in the years 1996-2000 and 2007, each year 
analyzed separately [@Tilman2001; @Tilman2006]. 
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. See Appendix \ref{SM-Data_details} for additional data details.

@@

<!--summary table with info of all data used in Paper-->
```{r tab_summary_data_info, echo=F, results="asis",message=F}
library(kableExtra)
library(dplyr)
tab_data_info<-as.data.frame(matrix(NA,nrow=8,ncol=4))
colnames(tab_data_info)<-c("Data","Category","No. of measurements we used", "References")
tab_data_info$Data<-c("Soil C and N","Bird body masses and BMR", "Mammal body masses and BMR", 
                      "Green spruce and other aphid species abundances", "Leaf-curling plum and other aphid first flight dates","\\textit{Ceratium furca} and other plankton taxa abundances", "Plant diversity and aboveground biomass","Methane-flux")
tab_data_info$Category<-c("Environmental","Species-trait","Species-trait",
                          "Population","Phenological","Population","Community level","Ecosystem functioning")
tab_data_info$`No. of measurements we used`<-c(paste(nrow(d_soilCN)," locations",sep=""),
                                               paste(nrow(d_birdsBMR)," birds",sep=""),
                                               paste(nrow(d_mammalsBMR)," mammals",sep=""),
                                               "10 locations with at least 30-yr. timeseries for each of 20 sp.",
                                               "10 locations with at least 30-yr. timeseries for each of 20 sp.",
                                               "14 locations with at least 45-yr. timeseries for each of 22 taxa",
                                               paste(nrow(d_cc[[5]])," plots",sep=""),
                                               "13 locations with at least 50 dates of data")
tab_data_info$References<-c("US Rapid Carbon Assessment database (RaCA), Wills et al. (2014)", "McNab (2009)","McNab (2008)",
                            "Rothamsted Insect Survey","Rothamsted Insect Survey","Continuous Plankton Recorder Survey",
                            "Cedar Creek Ecosystem Science Reserve, biodiversity experiment e120",
                            "Great Miami Wetland Mitigation Bank, Smyth et al. (2019)")

@@

In addition to our model selection approach, we also used 
nonparametric approaches to provide greater 
confidence in our answers to Q1. We used three 
statistics which quantify the extent to which 
the normalized ranks $u_i$ and $v_i$ are 
related in any part of their distributions; we here describe the 
statistics conceptually, with details in Appendix \ref{SM-nonparam_stats}.
The statistics are defined with positively associated
variables in mind. All our variables were positively associated
when they were significantly associated.
Given two bounds 
$0 \leq l_b < u_b \leq 1$, define the lines $u+v=2l_b$ and 
$u+v=2u_b$, which intersect the unit square (Figs \ref{fig_cop_pedag1}H, 
\ref{SM-fig_stat_image}). 
Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and $u_b\leq 0.5$ gives information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ gives information about dependence in 
the right parts. The first statistic, 
$\cor_{l_b,u_b}(u,v)$, is the portion of the Spearman 
correlation of $u_i$ and $v_i$ that is attributable to the points 
between the bounds.
Larger values indicate stronger positive association.
We also defined a statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar 
interpretation to $\cor_{l_{b},u_{b}}(u,v)$.
Our third statistic, $\Dsq_{l_b,u_b}(u,v)$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, for which large values inidicate strong dependence
between the bounds, small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6.5in]{./Results/BivarDataPlot.pdf}
\caption{Upper panels show raw data plots for (A) $\log_{10}$(soil C) and $\log_{10}$(soil N) data, (B) $\log_{10}$(basal metabolic rate (BMR)) vs. $\log_{10}$(body mass) 
for birds, (C) the same for mammals, and (D) above-ground plant biomass vs. Shannon's index (\textit{H}) from Cedar Creek. Bottom panels show the corresponding copula plots. See Data section for units used in upper panels. \label{fig_biv_multi_raw_cop}}
\end{center}
\end{figure}

@@

<!--bivariate example-->
For the soil C and N data (Data section, Table \ref{tab_data_info}, 
Fig. \ref{fig_biv_multi_raw_cop}A), variables were 
non-independent ($p=$ `r BivMS_res_Loecke$IndepTestRes`, to within the precision available 
from `BiCopIndTest` in the `VineCopula` package); non-independence is also visually apparent.
The Kendall correlation was `r round(BivMS_res_Loecke$TauVal,2)`. We fitted our 16 copulas
to the normalized ranks (Fig. \ref{fig_biv_multi_raw_cop}E) 
and computed AIC and BIC weights and corresponding
lower- and upper-tail statistics for the fitted distributions (Table \ref{SM-tab_soilCNfit}).
`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]` was the 
minimum-AIC copula, with AIC weight `r round(max(BivMS_res_Loecke$InfCritRes$AICw),2)`, whereas the 
N copula had AIC weight essentially 0, answering the 
first part of Q1 for these data. 
Model-averaged lower- and upper-tail dependence statistics were `r round(BivMS_res_Loecke$relLTdep_AICw,3)` 
and `r round(BivMS_res_Loecke$relUTdep_AICw,3)`, respectively (AIC weights were used for averaging). 
These values are distinct from what an N copula would give, namely, 0, answering the second part 
Q1 for these data. The values also differed substantially 
from each other, answering the third part of Q1. 
The numbers reflect stronger upper- than lower-tail
dependence, and this is also visible in the extreme upper-right
and lower-left corners of the copula plot (Fig. \ref{fig_biv_multi_raw_cop}E).

Although these results convincingly show non-N copula structure, 
we caution that the model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula 
(`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]`) 
was a poor fit ($p=$ `r BivMS_res_Loecke$GofRes_CvM` and `r BivMS_res_Loecke$GofRes_KS` for the 
Cramer-von Mises and Kolmogorov-Smirnov goodness-of-fit tests, respectively, to within the precision 
available from `BiCopGofTest`). However, nonparametric statistics verified
both that tail dependence was distinct from that of an N copula, and that it 
was asymmetric (Appendices \ref{SM-Q1bivarnonparam}, \ref{SM-surrog_test}, 
Fig. \ref{SM-fig_soilCN_nonparam}, Table \ref{SM-tab_soilCN_asym}). Our 
affirmative, empirically based answer to Q1 for the soil C and N data is 
represented in Fig. \ref{MasterFigure} as a solid outline around "soil C and N"
in the right-most box in the middle row.

<!--bivariate summary-->
Table \ref{tab_bivar_summary} summarizes Q1 results 
for the bivariate datasets. The two 
variables were significantly related for all datasets 
(Table \ref{tab_bivar_summary}, row 1). A non-N copula 
always emerged as the lowest-AIC (highest-AIC-weight) copula 
(Table \ref{tab_bivar_summary}, row 2). 
The N copula was always a poor fit compared to the highest-AIC-weight copula 
(Table \ref{tab_bivar_summary}, rows 3-4), answering the first
part of Q1. 
Lower- and upper-tail dependence statistics often differed substantially from 0 
(Table \ref{tab_bivar_summary}, rows 7-8) and from each other (Table \ref{tab_bivar_summary}, row 9), 
answering the second and third parts of Q1. 
The first three datasets had stronger upper- 
than lower-tail dependence, and the last dataset had the reverse, though weakly. 
Although these results convincingly show non-N copula structure, the 
model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula was a poor fit in some cases 
(Table \ref{tab_bivar_summary}, rows 5-6).
But our nonparametric statistics confirmed the same conclusions 
(Appendices \ref{SM-Q1bivarnonparam}, \ref{SM-surrog_test}).
Our 
affirmative, empirically based answer to Q1 for the bivariate datasets is 
represented in Fig. \ref{MasterFigure} as solid outlines around the names of those
datasets, in the middle row of boxes.
We did the same analysis 
with Cedar Creek data from other available years, 1996-1999 and 2007. For 1996 and 
2007, biomass and Shannon's index were independent; for 1997-1999, results 
were similar to 2000: the best-fitting copula was not N, and there was slightly stronger lower- 
than upper-tail dependence. Detailed results for each dataset are in 
Tables \ref{SM-tab_soilCNfit} - \ref{SM-tab_cc_asym_2000} and 
Figs. \ref{SM-fig_soilCN_nonparam} - \ref{SM-fig_cc2000_nonparam}.

We present green spruce aphid abundance (Table \ref{tab_data_info}) results
as an example multivariate analysis.
Independence was rejected for each pair of the 10 sampling locations. 
Best-fitting (lowest-AIC) copulas were non-N
for the large majority of location pairs (Table \ref{SM-tab_count_bestcop}), and 
AIC weights for best-fitting copulas (Table \ref{SM-tab_count_AICw}) 
were typically substantially 
larger than N-copula weights (Table \ref{SM-tab_count_AICw_normal}), answering the first part of Q1. Goodness of fit tests 
(Tables \ref{SM-tab_count_p_CvM}, \ref{SM-tab_count_p_KS}) in every case 
failed to reject the hypothesis that the AIC-best copula family was also 
an objectively adequate description of 
the data; i.e., the collection of 
copula families we used was sufficiently broad to characterize these data.
Model-averaged lower- (Table \ref{SM-tab_count_LTdep_AICw}) and upper-tail
(Table \ref{SM-tab_count_UTdep_AICw}) dependence statistics were exclusively 
positive and were frequently large, thereby differing 
from what an N copula would give (i.e., 0), and answering the second
part of Q1: these data have greater tail dependence 
(lower and upper) than would be expected from an N-copula null hypothesis. 
Model-averaged lower- minus upper-tail dependence statistics 
(Table \ref{SM-tab_count_LTmUT}) were positive 
for all but a few location pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`). Thus the spatial synchrony of rarity in the green spruce aphid is stronger 
than the spatial synchrony of outbreaks. This also answers the third part of Q1.
Nonparametric statistics verified that tail dependence was asymmetric for the green 
spruce aphid abundance data (Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}).

<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes Q1 results for
multivariate data. Results supported the 
conclusions that non-N copula structure, non-N tail dependence, and asymmetric tail 
dependence were common, answering Q1 in the affirmative for these data. 
Most location pairs were non-independent 
(Table \ref{tab_multivar_summary}, row 2). The large majority of non-independent location 
pairs had best-fitting copulas that were not the N copula (Table \ref{tab_multivar_summary}, 
row 3), and AIC weights of best-fitting copulas were 
better than those of N copulas (Table \ref{tab_multivar_summary}, rows 4-5).
Best-fitting copulas were nearly always considered an adequate fit 
(Table \ref{tab_multivar_summary}, row 6). Some of the datasets (green spruce aphid
abundance, \textit{Ceratium furca} abundance, methane-flux) showed stronger 
or marginally stronger lower- than upper-tail dependence 
(Table \ref{tab_multivar_summary}, rows 7-9), whereas leaf-curling plum aphid first flight
data showed the reverse. Asymmetry results were verified by nonparametric approaches 
<!--(e.g., Table \ref{tab_multivar_summary}, row 10)-->
(Appendix \ref{SM-Q1multvarnonparam}, \ref{SM-spatial_resamp}). Detailed results are in 
Tables \ref{SM-tab_count_bestcop}-\ref{SM-tab_methane_resamp} and 
Figs \ref{SM-fig_aphid_count}-\ref{SM-fig_methane}.
Our 
affirmative, empirically based answer to Q1 for the multivariate datasets is 
represented in Fig. \ref{MasterFigure} as solid outlines around the names
of those datasets, in the middle row of boxes.
We also carried out the same analyses for abundance and first-flight data for the 18 
aphid species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other than
\textit{Ceratium furca} (results not shown); results supported the conclusion that 
non-N copula structure and tail dependence, and asymmetry of tail dependence are 
common.

@@

\noindent \textbf{Figure 3.} Upper panels show raw data plots for (A) $\log_{10}$(soil C) and $\log_{10}$(soil N) data, (B) $\log_{10}$(basal metabolic rate (BMR)) vs. $\log_{10}$(body mass) 
for birds, (C) the same for mammals, and (D) above-ground plant biomass vs. Shannon's index (\textit{H}) from Cedar Creek. Bottom panels show the corresponding copula plots. See Data section for units used in upper panels.
